New cardinality estimation algorithms for HyperLogLog sketches

### 2016-11-21 paper/paper.tex
minor changes

% SYMBOLS
\newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}
\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symIndexJ}{j}
\symDefine{\symCount}{c}
\symDefine{\symCountVariate}{C}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symKappa}{\kappa}
\symDefine{\symX}{x}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symPhi}{\varphi}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symInteger}{z}
\symDefine{\symIntegerVariate}{Z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityRawEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symExpectation}{\mathbb{E}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symHelperW}{w}
\symDefine{\symHelperV}{v}
\symDefine{\symHelperU}{u}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}
\symDefine{\symPowerSeriesFunc}{\xi}
\symDefine{\symSmallCorrectionFunc}{\sigma}
\symDefine{\symLargeCorrectionFunc}{\tau}
\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}
\symDefine{\symBiasCorrectionFunc}{w_{\text{corr}}}

@@

Let us assume that the register values have following cumulative distribution function
\begin{equation}
\label{equ:assumed_register_val_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
\end{equation}
For now we ignore that this distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$. For a random variable $\symRegValVariate$ obeying \eqref{equ:assumed_register_val_distribution} the expectation of $2^{-\symRegValVariate}$ is given by
\begin{multline}
\label{equ:expectation_power_two}
\symExpectation(2^{-\symRegValVariate})
=
\\
\sum_{\symRegVal = -\infty}^\infty
2^{-\symRegVal}
\left(
e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}
-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal-1}}}
\right)
=
\frac{1}{2}
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
=
\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left(\symPoissonRate/\symNumReg\right)\right)}{\symPoissonRate},
\end{multline}
where the function 
\begin{equation}
\label{equ:power_series_function}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
 is a smooth periodic function with period 1. Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=\num{9.885e-6}$ (see \cref{fig:power_series_func}).

@@

Assume two given  HyperLogLog sketches with register values $\boldsymbol{\symRegValVariate}_1$ and $\boldsymbol{\symRegValVariate}_2$ representing the sets $\symSetS_1$ and $\symSetS_2$, respectively. The goal is to find estimates for the cardinalities of the pair-wise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$. The Poisson approximation allows us to assume that the elements of $\symSetA$ are inserted into the HyperLogLog sketch representing $\symSetS_1$ at rate $\symPoissonRate_\symSetASuffix$. Similarly, the elements of $\symSetB$ are inserted into the HyperLogLog sketch representing $\symSetS_2$ at rate $\symPoissonRate_\symSetBSuffix$. Furthermore, we assume that elements of $\symSetX$ are inserted into both HyperLogLog sketches simultaneously at rate $\symPoissonRate_\symSetXSuffix$. We expect that good estimates $\symPoissonRateEstimate_\symSetASuffix$, $\symPoissonRateEstimate_\symSetBSuffix$, and $\symPoissonRateEstimate_\symSetXSuffix$ for the rates are also good estimates for the cardinalities of $\symSetA$, $\symSetB$, and $\symSetX$.

\subsection{Joint log-likelihood funtion}
In order to get maximum-likelihood estimators for $\symPoissonRateEstimate_\symSetASuffix$, $\symPoissonRateEstimate_\symSetBSuffix$, and $\symPoissonRateEstimate_\symSetXSuffix$ we need to derive the joint probability distribution of two HyperLogLog sketches. Under the Poisson model the register values are independent and identically distributed. Therefore, we first derive the joint probability distribution for a single register that has value $\symRegValVariate_1$ in the first HyperLogLog sketch representing $\symSetS_1$ and value $\symRegValVariate_2$ in the second HyperLogLog sketch representing $\symSetS_2$.

The HyperLogLog sketch that represents $\symSetS_1$ can be thought to be constructed from two HyperLogLog sketches representing $\symSetA$ and $\symSetX$ and merging both using \eqref{equ:hll_union}. Analogously, the HyperLogLog sketch for $\symSetS_2$ could have been obtained from sketches for $\symSetB$ and $\symSetX$. Let  $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ be the value of the considered register in the HyperLogLog sketch representing $\symSetA$, $\symSetB$, and $\symSetX$, respectively. The corresponding values in sketches for $\symSetS_1$ and $\symSetS_2$ are given by
\begin{equation}
\symRegValVariate_1 = \max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right)
,\quad
\symRegValVariate_2 = \max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right).
\end{equation}
Their joint cumulative probability function is given as
\begin{align}
\symProbability\!\left(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
\right)
&=
\symProbability\!\left(
\max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_1
\wedge
\max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_2
\right)
\nonumber\\
&=
\symProbability
\!\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\wedge
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\wedge
\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
\right)
\nonumber\\
&=
\symProbability
\!\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\right)
\symProbability
\!\left(
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\right)
\symProbability
\!\left(
\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
\right).
\end{align}
Here the last transformation used the independence of $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$, because by definition, the sets $\symSetA$, $\symSetB$, and $\symSetX$ are disjoint. Furthermore, under the Poisson model $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ obey 
\eqref{equ:register_value_distribution}. If we take into account that elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetXSuffix$, $\symPoissonRate_\symSetASuffix$, and $\symPoissonRate_\symSetBSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
\begin{equation}
\symProbability(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
)
=\begin{cases}
0 & \symRegVal_1 < 0 \vee \symRegVal_2 < 0
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix}}{\symNumReg 2^{\symRegVal_1}}
-
\frac{\symPoissonRate_{\symSetBSuffix}}{\symNumReg 2^{\symRegVal_2}}
-
\frac{\symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\min(\symRegVal_1, \symRegVal_2)}}
}
& 0\leq\symRegVal_1 \leq \symRegRange \wedge 0\leq\symRegVal_2\leq\symRegRange
\\
e^{
-
\frac{\symPoissonRate_{\symSetBSuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_2}}
}
& 0\leq\symRegVal_2 \leq \symRegRange < \symRegVal_1
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_1}}
}
&  0\leq\symRegVal_1 \leq \symRegRange < \symRegVal_2
\\
1
&
\symRegRange < \symRegVal_1 \wedge \symRegRange < \symRegVal_2
\end{cases}
\end{equation}

@@

The log-likelihood function \eqref{equ:log_likelihood_pair} does not always have a strict global maximum point. For example, if all register values of the first HyperLogLog sketch are larger than the corresponding values in the second HyperLogLog sketch, that is $\symCountVariate_{1\symRegVal}^{<}
=
\symCountVariate_{\symRegVal}^{=}
=
\symCountVariate_{2\symRegVal}^{>}
=
0$ for all $\symRegVal$, the function can be rewritten as sum of two functions, one dependent on $\symPoissonRate_\symSetASuffix$ and the other dependent on $\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix$. 
The maximum is obtained, if $\symPoissonRate_\symSetASuffix = \symPoissonRateEstimate_1$ and $\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix = \symPoissonRateEstimate_2$. Here $\symPoissonRateEstimate_1$ and $\symPoissonRateEstimate_2$ are the maximum likelihood cardinality estimates for the first and second HyperLogLog sketch, respectively. This means that the maximum-likelihood makes no clear statement about the intersection size in this case. The estimate for $\symPoissonRate_\symX$ could be anything between 0 and $\symPoissonRateEstimate_2$. For comparison, the inclusion-exclusion approach would give $\symPoissonRateEstimate_2$ as estimate. This result is questionable, because there is no evidence in this case, that the sets $\symSetS_1$ and $\symSetS_2$ really have common elements.

@@

%$
%\symPoissonRate_\symSetASuffix
%=
%\symPoissonRate_\symSetBSuffix=0
%$ and all the register values must be pair-wise identical, which finally 
%yields \eqref{equ:log_likelihood_single} for $\symPoissonRate_\symSetXSuffix$.

The inclusion-exclusion method does not use all the available information given by the sufficient statistic \eqref{equ:sufficient_joint_statistic}, because the estimator is a function of the three vectors $(\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1})$, 
$(\boldsymbol\symCountVariate^{<}_{2} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2})$, and 
$(\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2})$. In contrast, the maximum-likelihood method incorporates each value of the sufficient statistic \eqref{equ:sufficient_joint_statistic} individually.

@@

%As demonstrated by we In all other cases, the maximum-likelihood estimates for $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$ will be non-zero. We use the conventional approach to get start values for the estimates of $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$, respectively. However, in order to ensure that the values are positive we restrict them to values greater than or equal to 1.

@@

\cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, and \cref{tbl:joint_estimation_x} compare the relative estimation error to the conventional approach using single sketch cardinality estimation together with \eqref{equ:conventional_approach}. The mean, the standard deviation, and the root-mean-square error are given. Furthermore, we calculated an improvement factor which represents the root-mean-square error ratio between both approaches. Since we only observed values larger than 1, the new maximum-likelihood estimation approach gives better estimates for all investigated cases. For some cases the improvement factor is clearly larger than 2. Due to square root error scaling law, this means that we would need 4 times more registers to get the same error when using the conventional approach. We also considered some cases with true cardinalities smaller  than the number of registers. As the results suggest and as expected the joint estimation algorithm works well over the entire cardinality range without special handling of small cardinalities.

@@

Using the maximum-likelihood method we have been able to improve the cardinality estimates for the results of set operations between two HyperLogLog sketches. Unfortunately, joint cardinality estimation is much more expensive than for a single HyperLogLog sketch, because it requires maximization of a multi-dimensional function. Since we have found the corrected raw estimator which is almost as accurate as the maximum-likelihood estimator for the single HyperLogLog case, we could imagine that there also exists a faster algorithm for the two HyperLogLog case. It is expected that such a new algorithm makes use of all the information given by the sufficient statistic \eqref{equ:sufficient_joint_statistic}.

The presented maximum-likelihood method could also be used to estimate the cardinality of set operations between more than two HyperLogLog sketches. However, further research is necessary to determine, if this is feasible from a practical point of view. As for the inclusion-exclusion principle, the effort would scale exponentially with the number of involved HyperLogLog sketches.

The maximum-likelihood method can also be used to estimate similarity measures such as the Jaccard index of two sets that are represented as HyperLogLog sketches. This directly leads to the question whether the HyperLogLog algorithm could be used for locality-sensitive hashing \cite{Leskovec2014, Wang2014}. Various locality-sensitive hashing algorithms have been proposed in the past. Among the most popular ones are the SimHash \cite{Charikar2002} and the min-wise hashing \cite{Broder1997, Li2011} algorithms whose hash collision probabilities are a function of the cosine and Jaccard similarities, respectively.

@@

We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches based on the Poisson approximation. For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation. The first uses the original estimator extended by theoretically motivated correction terms. The second is based on the maximum-likelihood method and solves the corresponding optimization problem using the secant method. Both algorithms are inherently unbiased and can be implemented without dependence on empirically determined bias correction data.

>>>>

% SYMBOLS
\newcommand*{\symDefine}[2]{\newcommand{{#1}}{{#2}}}
\symDefine{\symPrecision}{p}
\symDefine{\symRegRange}{q}
\symDefine{\symRegVal}{k}
\symDefine{\symRegValVariate}{K}
\symDefine{\symNumReg}{m}
\symDefine{\symDataItem}{D}
\symDefine{\symBitRepA}{a}
\symDefine{\symBitRepB}{b}
\symDefine{\symIndexI}{i}
\symDefine{\symIndexJ}{j}
\symDefine{\symIndexL}{l}
\symDefine{\symCount}{c}
\symDefine{\symCountVariate}{C}
\symDefine{\symAlpha}{\alpha}
\symDefine{\symBeta}{\beta}
\symDefine{\symA}{a}
\symDefine{\symB}{b}
\symDefine{\symKappa}{\kappa}
\symDefine{\symX}{x}
\symDefine{\symY}{y}
\symDefine{\symZ}{z}
\symDefine{\symPhi}{\varphi}
\symDefine{\symXEstimate}{\hat{\symX}}
\symDefine{\symXNormalized}{\symX'}
\symDefine{\symN}{n}
\symDefine{\symS}{s}
\symDefine{\symInteger}{z}
\symDefine{\symIntegerVariate}{Z}
\symDefine{\symCardinality}{n}
\symDefine{\symCardinalityEstimate}{\hat{\symCardinality}}
\symDefine{\symCardinalityRawEstimate}{\symCardinalityEstimate_\text{raw}}
\symDefine{\symError}{\varepsilon}
\symDefine{\symStopDelta}{\delta}
\symDefine{\symStopEpsilon}{\varepsilon}
\symDefine{\symBigO}{\mathcal{O}}
\symDefine{\symExpectation}{\mathbb{E}}
\symDefine{\symProbability}{P}
\symDefine{\symProbabilityMass}{\rho}
\symDefine{\symRegProbability}{\gamma}
\symDefine{\symLikelihood}{\mathcal{L}}
\symDefine{\symPoissonRate}{\lambda}
\symDefine{\symPoissonRateEstimate}{\hat{\symPoissonRate}}
\symDefine{\symFunc}{f}
\symDefine{\symFuncPrime}{g}
\symDefine{\symHelper}{h}
\symDefine{\symHelperApprox}{\tilde{\symHelper}}
\symDefine{\symHelperW}{w}
\symDefine{\symHelperV}{v}
\symDefine{\symHelperU}{u}
\symDefine{\symSetA}{A}
\symDefine{\symSetB}{B}
\symDefine{\symSetS}{S}
\symDefine{\symSetX}{X}
\symDefine{\symSetASuffix}{a}
\symDefine{\symSetBSuffix}{b}
\symDefine{\symSetXSuffix}{x}
\symDefine{\symCountMatrix}{\mathbf{c}}
\symDefine{\symPowerSeriesFunc}{\xi}
\symDefine{\symSmallCorrectionFunc}{\sigma}
\symDefine{\symLargeCorrectionFunc}{\tau}
\symDefine{\symEpsPowerSeriesFunc}{\varepsilon_\symPowerSeriesFunc}
\symDefine{\symBiasCorrectionFunc}{w_{\text{corr}}}

@@

Let us assume that the register values have following cumulative distribution function
\begin{equation}
\label{equ:assumed_register_val_distribution}
\symProbability(\symRegValVariate \leq \symRegVal\vert\symPoissonRate) = e^{-\frac{\symPoissonRate}{\symNumReg 2^{ \symRegVal}}}.
\end{equation}
For now we ignore that this distribution has infinite support and differs from the register value distribution under the Poisson model \eqref{equ:register_value_distribution}, whose support is limited to the range $[0, \symRegRange+1]$. For a random variable $\symRegValVariate$ obeying \eqref{equ:assumed_register_val_distribution} the expectation of $2^{-\symRegValVariate}$ is given by
\begin{multline}
\label{equ:expectation_power_two}
\symExpectation(2^{-\symRegValVariate})
=
\\
\sum_{\symRegVal = -\infty}^\infty
2^{-\symRegVal}
\left(
e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal}}}
-e^{-\frac{\symPoissonRate}{\symNumReg 2^{\symRegVal-1}}}
\right)
=
\frac{1}{2}
\sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal}
e^{-\frac{\symPoissonRate}{\symNumReg} 2^{\symRegVal}}
=
\frac{\symAlpha_\infty\,\symNumReg\,\symPowerSeriesFunc\!\left(\log_2\!\left(\symPoissonRate/\symNumReg\right)\right)}{\symPoissonRate},
\end{multline}
where the function 
\begin{equation}
\label{equ:power_series_function}
\symPowerSeriesFunc(\symX):= \log(2) \sum_{\symRegVal = -\infty}^\infty
2^{\symRegVal+\symX}
e^{-2^{\symRegVal+\symX}}
\end{equation}
is a smooth periodic function with period 1. 
%Its Fourier series is
%\begin{equation}
%\symPowerSeriesFunc(\symX) = 1 + \sum_{\symIndexL=1}^{\infty}
%\symA_\symIndexL \cos(2\pi\symIndexL\symX) + \symB_\symIndexL \sin(2\pi\symIndexL\symX)
%\end{equation}
%with coefficients
%\begin{equation}
%\begin{aligned}
%\symA_\symIndexL &= 2\log(2)\int_{-\infty}^\infty 2^\symX e^{-2^\symX}\cos(2\pi\symIndexL\symX)d\symX, \\
%\symB_\symIndexL &= 2\log(2)\int_{-\infty}^\infty 2^\symX e^{-2^\symX}\sin(2\pi\symIndexL\symX)d\symX.
%\end{aligned}
%\end{equation}
%Numerical evaluations give for the $\symA_1\approx \num{6.353e-6}$ and $\symB_1\approx \num{-7.572e-6}$.
Numerical evaluations indicate that this function can be bounded by $1 - \symEpsPowerSeriesFunc 
\leq \symPowerSeriesFunc(\symX) \leq 1 + \symEpsPowerSeriesFunc$ with $\symEpsPowerSeriesFunc:=\num{9.885e-6}$ (see \cref{fig:power_series_func}).

@@

Assume two given  HyperLogLog sketches with register values $\boldsymbol{\symRegValVariate}_1$ and $\boldsymbol{\symRegValVariate}_2$ representing the sets $\symSetS_1$ and $\symSetS_2$, respectively. The goal is to find estimates for the cardinalities of the pairwise disjoint sets $\symSetX = \symSetS_1\cap\symSetS_2$, $\symSetA = \symSetS_1\setminus\symSetS_2$, and $\symSetB = \symSetS_2\setminus\symSetS_1$. The Poisson approximation allows us to assume that the elements of $\symSetA$ are inserted into the HyperLogLog sketch representing $\symSetS_1$ at rate $\symPoissonRate_\symSetASuffix$. Similarly, the elements of $\symSetB$ are inserted into the HyperLogLog sketch representing $\symSetS_2$ at rate $\symPoissonRate_\symSetBSuffix$. Furthermore, we assume that elements of $\symSetX$ are inserted into both HyperLogLog sketches simultaneously at rate $\symPoissonRate_\symSetXSuffix$. We expect that good estimates $\symPoissonRateEstimate_\symSetASuffix$, $\symPoissonRateEstimate_\symSetBSuffix$, and $\symPoissonRateEstimate_\symSetXSuffix$ for the rates are also good estimates for the cardinalities of $\symSetA$, $\symSetB$, and $\symSetX$.

\subsection{Joint log-likelihood funtion}
In order to get maximum likelihood estimators for $\symPoissonRateEstimate_\symSetASuffix$, $\symPoissonRateEstimate_\symSetBSuffix$, and $\symPoissonRateEstimate_\symSetXSuffix$ we need to derive the joint probability distribution of two HyperLogLog sketches. Under the Poisson model the register values are independent and identically distributed. Therefore, we first derive the joint probability distribution for a single register that has value $\symRegValVariate_1$ in the first HyperLogLog sketch representing $\symSetS_1$ and value $\symRegValVariate_2$ in the second HyperLogLog sketch representing $\symSetS_2$.

The HyperLogLog sketch that represents $\symSetS_1$ can be thought to be constructed from two HyperLogLog sketches representing $\symSetA$ and $\symSetX$ and merging both using \eqref{equ:hll_union}. Analogously, the HyperLogLog sketch for $\symSetS_2$ could have been obtained from sketches for $\symSetB$ and $\symSetX$. Let  $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ be the value of the considered register in the HyperLogLog sketch representing $\symSetA$, $\symSetB$, and $\symSetX$, respectively. The corresponding values in sketches for $\symSetS_1$ and $\symSetS_2$ are given by
\begin{equation}
\symRegValVariate_1 = \max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right)
,\quad
\symRegValVariate_2 = \max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right).
\end{equation}
Their joint cumulative probability function is given as
\begin{align}
\symProbability\!\left(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
\right)
&=
\symProbability\!\left(
\max\!\left(\symRegValVariate_\symSetASuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_1
\wedge
\max\!\left(\symRegValVariate_\symSetBSuffix, \symRegValVariate_\symSetXSuffix\right) \leq \symRegVal_2
\right)
\nonumber\\
&=
\symProbability
\!\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\wedge
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\wedge
\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
\right)
\nonumber\\
&=
\symProbability
\!\left(
\symRegValVariate_\symSetASuffix \leq \symRegVal_1
\right)
\symProbability
\!\left(
\symRegValVariate_\symSetBSuffix \leq \symRegVal_2
\right)
\symProbability
\!\left(
\symRegValVariate_\symSetXSuffix \leq \min\!\left(\symRegVal_1, \symRegVal_2\right)
\right).
\end{align}
Here the last transformation used the independence of $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$, because by definition, the sets $\symSetA$, $\symSetB$, and $\symSetX$ are pairwise disjoint. Furthermore, under the Poisson model $\symRegValVariate_\symSetASuffix$, $\symRegValVariate_\symSetBSuffix$, and $\symRegValVariate_\symSetXSuffix$ obey 
\eqref{equ:register_value_distribution}. If we take into account that elements are added to $\symSetA$, $\symSetB$, and $\symSetX$ at rates $\symPoissonRate_\symSetXSuffix$, $\symPoissonRate_\symSetASuffix$, and $\symPoissonRate_\symSetBSuffix$, respectively, the probability that a certain register has a value less than or equal to $\symRegVal_1$ in the first HyperLogLog sketch and simultaneously a value less than or equal to $\symRegVal_2$ in the second one can be written as
\begin{equation}
\symProbability(
\symRegValVariate_1 \leq \symRegVal_1
\wedge
\symRegValVariate_2 \leq \symRegVal_2
)
=\begin{cases}
0 & \symRegVal_1 < 0 \vee \symRegVal_2 < 0
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix}}{\symNumReg 2^{\symRegVal_1}}
-
\frac{\symPoissonRate_{\symSetBSuffix}}{\symNumReg 2^{\symRegVal_2}}
-
\frac{\symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\min(\symRegVal_1, \symRegVal_2)}}
}
& 0\leq\symRegVal_1 \leq \symRegRange \wedge 0\leq\symRegVal_2\leq\symRegRange
\\
e^{
-
\frac{\symPoissonRate_{\symSetBSuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_2}}
}
& 0\leq\symRegVal_2 \leq \symRegRange < \symRegVal_1
\\
e^{
-
\frac{\symPoissonRate_{\symSetASuffix} + \symPoissonRate_{\symSetXSuffix}}{\symNumReg 2^{\symRegVal_1}}
}
&  0\leq\symRegVal_1 \leq \symRegRange < \symRegVal_2
\\
1
&
\symRegRange < \symRegVal_1 \wedge \symRegRange < \symRegVal_2
\end{cases}
\end{equation}

@@

The log-likelihood function \eqref{equ:log_likelihood_pair} does not always have a strict global maximum point. For example, if all register values of the first HyperLogLog sketch are larger than the corresponding values in the second HyperLogLog sketch, that is $\symCountVariate_{1\symRegVal}^{<}
=
\symCountVariate_{\symRegVal}^{=}
=
\symCountVariate_{2\symRegVal}^{>}
=
0$ for all $\symRegVal$, the function can be rewritten as sum of two functions, one dependent on $\symPoissonRate_\symSetASuffix$ and the other dependent on $\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix$. 
The maximum is obtained, if $\symPoissonRate_\symSetASuffix = \symPoissonRateEstimate_1$ and $\symPoissonRate_\symSetBSuffix+\symPoissonRate_\symSetXSuffix = \symPoissonRateEstimate_2$. Here $\symPoissonRateEstimate_1$ and $\symPoissonRateEstimate_2$ are the maximum likelihood cardinality estimates for the first and second HyperLogLog sketch, respectively. This means that the maximum likelihood method makes no clear statement about the intersection size in this case. The estimate for $\symPoissonRate_\symX$ could be anything between 0 and $\symPoissonRateEstimate_2$. For comparison, the inclusion-exclusion approach would give $\symPoissonRateEstimate_2$ as estimate. This result is questionable, because there is no evidence in this case, that the sets $\symSetS_1$ and $\symSetS_2$ really have common elements.

@@

%$
%\symPoissonRate_\symSetASuffix
%=
%\symPoissonRate_\symSetBSuffix=0
%$ and all the register values must be pairwise identical, which finally 
%yields \eqref{equ:log_likelihood_single} for $\symPoissonRate_\symSetXSuffix$.

The inclusion-exclusion method does not use all the available information given by the sufficient statistic \eqref{equ:sufficient_joint_statistic}, because the estimator is a function of the three vectors $(\boldsymbol\symCountVariate^{<}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{1})$, 
$(\boldsymbol\symCountVariate^{<}_{2} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2})$, and 
$(\boldsymbol\symCountVariate^{>}_{1} + \boldsymbol\symCountVariate^{=} + \boldsymbol\symCountVariate^{>}_{2})$. In contrast, the maximum likelihood method incorporates each value of the sufficient statistic \eqref{equ:sufficient_joint_statistic} individually.

@@

%As demonstrated by we In all other cases, the maximum likelihood estimates for $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$ will be non-zero. We use the conventional approach to get start values for the estimates of $\symPoissonRate_\symSetASuffix$, $\symPoissonRate_\symSetBSuffix$, and $\symPoissonRate_\symSetXSuffix$, respectively. However, in order to ensure that the values are positive we restrict them to values greater than or equal to 1.

@@

\cref{tbl:joint_estimation_a}, \cref{tbl:joint_estimation_b}, and \cref{tbl:joint_estimation_x} compare the relative estimation error to the conventional approach using single sketch cardinality estimation together with \eqref{equ:conventional_approach}. The mean, the standard deviation, and the root-mean-square error are given. Furthermore, we calculated an improvement factor which represents the root-mean-square error ratio between both approaches. Since we only observed values larger than 1, the new maximum likelihood estimation approach gives better estimates for all investigated cases. For some cases the improvement factor is clearly larger than 2. Due to square root error scaling law, this means that we would need 4 times more registers to get the same error when using the conventional approach. We also considered some cases with true cardinalities smaller  than the number of registers. As the results suggest and as expected the joint estimation algorithm works well over the entire cardinality range without special handling of small cardinalities.

@@

Using the maximum likelihood method we have been able to improve the cardinality estimates for the results of set operations between two HyperLogLog sketches. Unfortunately, joint cardinality estimation is much more expensive than for a single HyperLogLog sketch, because it requires maximization of a multi-dimensional function. Since we have found the corrected raw estimator which is almost as accurate as the maximum likelihood estimator for the single HyperLogLog case, we could imagine that there also exists a faster algorithm for the two HyperLogLog case. It is expected that such a new algorithm makes use of all the information given by the sufficient statistic \eqref{equ:sufficient_joint_statistic}.

The presented maximum likelihood method could also be used to estimate the cardinality of set operations between more than two HyperLogLog sketches. However, further research is necessary to determine, if this is feasible from a practical point of view. As for the inclusion-exclusion principle, the effort would scale exponentially with the number of involved HyperLogLog sketches.

The maximum likelihood method can also be used to estimate similarity measures such as the Jaccard index of two sets that are represented as HyperLogLog sketches. This directly leads to the question whether the HyperLogLog algorithm could be used for locality-sensitive hashing \cite{Leskovec2014, Wang2014}. Various locality-sensitive hashing algorithms have been proposed in the past. Among the most popular ones are the SimHash \cite{Charikar2002} and the min-wise hashing \cite{Broder1997, Li2011} algorithms whose hash collision probabilities are a function of the cosine and Jaccard similarities, respectively.

@@

We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches based on the Poisson approximation. For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation. The first uses the original estimator extended by theoretically motivated correction terms. The second is based on the maximum likelihood method and solves the corresponding optimization problem using the secant method. Both algorithms are inherently unbiased and can be implemented without dependence on empirically determined bias correction data.
